---
title: "Versioned, lineage-tracked storage for PhysioExperiment objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Versioned, lineage-tracked storage for PhysioExperiment objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Every lake chunk is gated on the optional OmicsLake backend so this vignette
# builds whether or not the backend is installed.
has_ol <- requireNamespace("OmicsLake", quietly = TRUE)
```

PhysioLake connects the PhysioExperiment ecosystem to
[OmicsLake](https://github.com/matsui-lab/OmicsLake) -- a versioned, on-disk
data-management engine (DuckDB + Apache Arrow + Parquet) with automatic
cross-dataset lineage. PhysioLake does **not** reimplement a data lake; it adds
a thin layer so that:

* a `PhysioExperiment` round-trips with **full fidelity** -- its class identity,
  `samplingRate` slot, non-2-D (epoched) assays, and W3C-PROV metadata all
  survive, and
* each object's **per-object (micro) provenance** becomes a first-class node in
  OmicsLake's **cross-dataset (macro) lineage**.

OmicsLake is an optional backend (a `Suggests` dependency), so PhysioLake stays
standalone-installable. The code below runs only when it is available:

```{r avail}
has_ol
```

## A tiny local lake

A `Lake`'s first constructor argument is a project *name*, not a path; the lake
lives under `root`. Here we create a throwaway lake in the session temp dir.

```{r lake, eval = has_ol}
library(PhysioLake)
lake <- OmicsLake::Lake$new(basename(tempfile("lake")), root = tempdir())
```

## Store and retrieve a PhysioExperiment

`physioPut()` stores the object (and, if present, its operation DAG as a
queryable companion table wired in as a lineage dependency). `lake$get()`
returns a genuine `PhysioExperiment`, not a bare `SummarizedExperiment`.

```{r roundtrip, eval = has_ol}
pe <- PhysioExperiment::PhysioExperiment(
  assays = list(raw = matrix(as.numeric(1:20), nrow = 10, ncol = 2)),
  samplingRate = 100
)
physioPut(lake, "subj01", pe)

restored <- lake$get("subj01")
class(restored)
PhysioExperiment::samplingRate(restored)
```

The sampling rate and class identity survive the round trip -- the piece the
generic SummarizedExperiment adapter would otherwise drop.

## Store a run bundle with internal lineage

`physioPutBundle()` stores the parts of a reproducible run (pre-registration,
results, reports, ...) as individual lake entries wired together with lineage
edges, so the run's internal structure is queryable.

```{r bundle, eval = has_ol}
components <- list(
  prereg = data.frame(key = "alpha", value = 1),
  result = data.frame(metric = "hr", value = 72)
)
physioPutBundle(lake, "run001", components,
                edges = list(result = "prereg"))

# `result` records `prereg` as an upstream dependency:
lake$deps("run001__result")
```

## Where to go next

* `?physioPut` / `?physioProvenance` -- object storage and provenance retrieval.
* `?physioPutBundle` -- storing a whole reproducible run.
* `?PhysioExperimentAdapter` -- the fidelity-preserving storage adapter.
* OmicsLake's own documentation for snapshots, time-travel, and lineage queries
  (`lake$snap()`, `lake$tree()`, `lake$impact()`).
